Won’t you be my neighboR?

### 2019-05-16 demo-doc.Rmd
update

@@

@@

@@

Try it out!

Select three *different* variables

@@

Try it out!

@@

@@

```{r}

>>>>

# Demo doc (part 1): Try it out!

Let's first load the tidyverse suite of packages.

If you haven't already, make sure to install the tidyverse, first.

*You can skip this code "chunk" if you already have the tidyverse installed; you only need to do this once*

```{r}
install.packages("tidyverse")
```

Be sure to run this next line to load the tidyverse. You do this each time you open R.

@@

Here is the data set we will use:

@@

Let's select only a few variables.

@@

**Try it out!**

Now, select three *different* variables, and run the code chunk to view the result.

@@

**Try it out!**

@@

Be sure to run the result to see whether it did what you think it should do!

@@

Let's create a new variable for those 60 years of age or older.

```{r}
gss_cat %>% 
  mutate(potential_senior = if_else(age >= 60, TRUE, FALSE))
```

### 2019-05-16 msu-workshop-2019.Rmd
update

# Who am I?

.pull-left[
* Assistant Professor: STEM Education, University of Tennessee, Knoxville
* Dad (one-year-old toddler!)
* Primary areas of interest
+ Data science in education (network analytic methods, experience sampling method, computational grounded theory)
+ Data science education (integrating data science and science education)
]

@@

@@

1. Why learn R?
1. Getting started with R Studio
1. Visualizing data
1. Processing data
1. Learning and doing more with R
1. Modeling data (time permitting)
1. Computer programming basics (time permitting)
1. Creating reproducible reports (time permitting)

@@

* It is capable of carrying out basic and complex statistical analyses
* It is able to work with data small (*n* = 30!) and large (*n* = 100,000+) efficiently
* It is a programming language and so is quite flexible
* There is a great, inclusive community of users and developers (and teachers)
* It is cross-platform, open-source, and freely-available

@@

# To download R Studio:

- Visit this page to download R studio: [https://www.rstudio.com/products/rstudio/download/](https://www.rstudio.com/products/rstudio/download/)
- Find your operating system (Mac, Windows, or Linux)
- Download the 'latest release' on the page for your operating system and download and install the application

If you do have issues, consider this [page](https://datacarpentry.org/R-ecology-lesson/), and then reach out for help. One good place to start is the [R Studio Community](https://community.rstudio.com/) is a great place to start.

### Check that it worked

@@

---
# Creating Projects

@@

@@

```{r, eval = FALSE}
install.packages("dplyr", repos = "http://cran.us.r-project.org")
```

You can also navigate to the Packages pane, and then click "Install", which will work the same as the line of code above. This is a way to install a package using code or part of the R Studio interface. Usually, writing code is a bit quicker, but using the interface can be very useful and complimentary to use of code.

@@

> a package is a like a book, a library is like a library; you use library() to check a package out of the library
> - Hadley Wickham, Chief Scientist, R Studio

@@

Once you have loaded the package in your session, you can run the functions that are contained within that package. To find a list of all those functions, you can run this in the R Studio console:

@@

```{r, eval = FALSE}
vignette(package = "dplyr")
vignette("dplyr", package = "dplyr")
```

@@

```{r, eval = FALSE}
??dplyr::filter
```

@@

```{r}
dat <- # example data frame
  data.frame(stringsAsFactors=FALSE,
             letter = c("A", "A", "A", "B", "B"),
             number = c(1L, 2L, 3L, 4L, 5L))

dat

filter(dat, letter == "A") # using dplyr::filter
```

@@

One important consideration is whether to save your workspace when you close R Studio. By default, R Studio saves all of the objects in your environment. This means that any data that you have loaded--or new data or objects that you have created, such as by merging two data sets together or creating a plot--will, by default, still exist when you open R Studio next. In general, this is not ideal, because it means that you may have taken steps interactively that are not documented your code. This means that when you share your code, or re-run it from the start, it may not work. An easy way to change this is to tell R Studio to start from scratch (in terms of your workspace) each time you open it. You can do that by changing the dropdown menu pointed out in the image below to "Never".

![optional caption text](images/save-workspace-reminder.jpg)
---

class: inverse, center, middle
# Processing data

@@

# The Tidyverse

The Tidyverse is a set of packages for data manipulation, exploration, and visualization using the design philosophy of 'tidy' data. Tidy data has a specific structure: each variable is a column, each observation is a row, and each type of observational unit is a table.

@@

**For more information on tidy data, check out [Hadley Wickhams's Tidy Data paper](http://vita.had.co.nz/papers/tidy-data.html).**
---

# Loading Data from Various Sources

### Saving a File from the Web

You'll need to copy this URL:

`https://goo.gl/bUeMhV`

Here's what it resolves to (it's a CSV file):

`https://raw.githubusercontent.com/data-edu/data-science-in-education/master/data/pisaUSA15/stu-quest.csv`

@@

student_responses_file_name <-
  paste0(getwd(), "./data/student-responses-data.csv")

@@

It may take a few seconds to download as it's around 20 MB.

The process above involves many core data science ideas and ideas from programming/coding. We will walk through them step-by-step.

@@

```{r}
student_responses_url <-
  "https://goo.gl/bUeMhV"
```

@@

2. We concatenate your working directory file path to the desired file name for the CSV using a *function* called `paste0`. This is stored in another *object* called `student_reponses_file_name`. This creates a file name with a *file path* in your working directory and it saves the file in the folder that you are working in.

```{r}
student_responses_file_name <-
  paste0(getwd(), "./data/student-responses-data.csv")
```

@@

Understanding how R is working in these terms can be helpful for troubleshooting and reaching out for help. It also helps you to use functions that you have never used before because you are familiar with how some functions work.

# Loading the file

Let's load the tidyverse library:

After loading the tidyverse packages, we can now load a file. We are going to call the data `student_responses`:

```{r, eval = TRUE}
student_responses <-
  read_csv("./data/student-responses-data.csv")
```

# Loading Excel files

We can then use the function `read_excel()` in the same way as `read_csv()`, where "path/to/file.xlsx" is where an Excel file you want to load is located (note that this code is not run here):

# Loading SAV files

# Saving Files

@@

We're also going to introduce a powerful, unusual *operator* in R, the pipe. The pipe is this symbol: `%>%`. It lets you *compose* functions. It does this by passing the output of one function to the next. A handy shortcut for writing out `%>%` is Command + Shift + M.

Here's an example. Let's say that we want to select a few variables from the `student_responses` dataset and save those variables into a new object, `student_mot_vars`. Here's how we would do that using `dplyr::select()`.

```{r}
student_mot_vars <- # save object student_mot_vars by...
  student_responses %>% # using dataframe student_responses
  select(SCIEEFF, JOYSCIE, INTBRSCI, EPIST, INSTSCIE) # and selecting only these five variables
```

@@

We can also rename the variables at the same time we select them. I put these on separate lines so I could add the comment, but you could do this all in the same line, too. It does not make a difference in terms of how `select()` will work.

```{r}
student_mot_vars <- # save object student_mot_vars by...
  student_responses %>% # using dataframe student_responses
  select(student_efficacy = SCIEEFF, # selecting variable SCIEEFF and renaming to student_efficiency
         student_joy = JOYSCIE, # selecting variable JOYSCIE and renaming to student_joy
         student_broad_interest = INTBRSCI, # selecting variable INTBRSCI and renaming to student_broad_interest
         student_epistemic_beliefs = EPIST, # selecting variable EPIST and renaming to student_epistemic_beliefs
         student_instrumental_motivation = INSTSCIE # selecting variable INSTSCIE and renaming to student_instrumental_motivation
  )
```

@@

# Processing data (part 1 of 4)

* `select()`
* `rename()`
* `filter()`
* `mutate()`

@@

# Processing data (part 2 of 4)

* `group_by()`
* `count()`
* `summarize()`

@@

# Processing data (part 3 of 4)

* `left_join()`
* `right_join()`
* `semi_join()`
* `anti_join()`

@@

# Processing data (part 4 of 4)

* `gather()`
* `spread()`

@@

class: inverse, center, middle
# Creating reproducible reports

@@

# Computer programming basics

# Objects and variables

```{r}
my_object <- 1
my_other_object <- c(1:10)
```

Reading in a CSV and assigning it to an object

```{r}
my_df <- read_csv("example-data.csv")
my_df
```

```{r}
library(tibble)
tibble(var1 = vec1,
       var2 = vec2)
```

# Functions

```{r}
mean(c(1, 2, 3))
```

```{r}
mean(c(1, 2, 3, NA))
```

```{r, eval = FALSE}
?mean
```

```{r}
mean(c(1, 2, 3, NA), na.rm = TRUE)
```

# Functions

```{r, eval = FALSE}
library(tidyverse)
read_directory <- function(directory) {
  map_df(list.files(directory), read_csv)
}

d <- read_directory("data")
```

# Iteration

```{r}
random_numbers <- rnorm(n = 10, mean = 0)
random_numbers
mean(random_numbers)
sd(random_numbers)
```

```{r}
library(tidyverse)
1:10 %>%
  map(~ rnorm(n = 10, mean = .x)) %>%
  map_dbl(mean)
```

>>>>

# #whoami1

.pull-left[
* Dr. Joshua Rosenberg
* Assistant Professor, STEM Education, University of Tennessee, Knoxville
* Dad (one-year-old toddler!)
* Primary areas of interest
  * Data science in education
  * Data science education
]

@@

# #whoami2

.pull-left[
* Dr. Emily Bovee
* Director of Educational Development and Assessment: Marquette University School of Dentistry
* Cat mom
* Primary areas of interest
  * Understanding student motivation in higher education and professional education
  * Data science education 
]

```{r, out.width = "350px", echo = FALSE, fig.align = "center"}
include_graphics("img/ebv.jpg")
```

@@

1. Why learn R?
1. Getting started with R Studio
1. Computer science foundations
1. Processing data
1. Loading data
1. Visualizing data
1. Modeling data 
1. Learning and doing more with R
1. Questions and sundry tools (time permitting)

@@

* It is capable of carrying out basic and complex statistical analyses
* It is able to work with data small (*n* = 30!) and large (*n* = 100,000+) efficiently
* It is a programming language and so is quite flexible
* There is a great, inclusive community of users and developers (and teachers)
* It is cross-platform, open-source, and freely-available
* It can h
---

@@

# Installations

To download R:
- Visit this page to download R: https://cran.r-project.org/
- Find your operating system (Mac, Windows, or Linux)
- Download the 'latest release' on the page for your operating system and download and install the application

To download R Studio:
- Visit this page to download R studio: https://www.rstudio.com/products/rstudio/download/
- Find your operating system (Mac, Windows, or Linux)
- Download the 'latest release' on the page for your operating system and download and install the application

## Check that it worked

@@

# Creating Projects

@@

Names should be short and easy-to-remember but also informative.

@@

```{r, eval = FALSE}
install.packages("dplyr")
```

You can also navigate to the Packages pane, and then click "Install", which will work the same as the line of code above. This is a way to install a package using code or part of the R Studio interface.

Usually, writing code is a bit quicker, but using the interface can be very useful and complimentary to use of code.

@@

> A package is a like a book, a library is like a library; you use library() to check a package out of the library
> - Hadley Wickham, Chief Scientist, R Studio

@@

Once you have loaded the package in your session, you can run the functions that are contained within that package.

To find a list of all those functions, you can run this in the R Studio console:

@@

Vignettes are long-form documentation (and tutorials) that can provide a helpful introduction to a package.

Run `vignette()` in order to view *all* of the vignettes available for a package:

```{r, eval = FALSE}
vignette(package = "dplyr")
```

Then, you can load a specific vignette.

```{r, eval = FALSE}
vignette("dplyr", package = "dplyr")
```

These are also available through CRAN (i.e., https://cran.r-project.org/web/packages/dplyr/index.html)
---

@@

```{r, eval = FALSE}
?dplyr::filter
```

@@

```{r}
dat <- # example data frame
  tibble(letter = c("A", "A", "A", "B", "B"),
         number = c(1L, 2L, 3L, 4L, 5L))

filter(dat, letter == "A")
```

@@

By default, R Studio saves all of the objects in your environment. In general, this is not ideal, because it means that you may have taken steps interactively that are not documented your code.

```{r, out.width = "425px", echo= FALSE, fig.align = "center"}
include_graphics("img/save-workspace-reminder.jpg")
```

@@

# The tidyverse

The tidyverse is a set of packages for data manipulation, exploration, and visualization using the design philosophy of 'tidy' data. Tidy data has a specific structure: each variable is a column, each observation is a row, and each type of observational unit is a table.

@@

# Computer science foundations
---

```{r}
tibble(var1 = vec1,
       var2 = vec2)
```

---
# Objects, functions, directories, and data structures

Here's a short URL: `https://goo.gl/bUeMhV`

This is what it resolves to (it's a CSV file): `https://raw.githubusercontent.com/data-edu/data-science-in-education/master/data/pisaUSA15/stu-quest.csv`

This chunk of code will save that file to a data directory in our working environment.

@@

student_responses_file_name <-
  paste0(getwd(), "/data/student-responses-data.csv")

@@

@@

```{r, eval = FALSE}
student_responses_url <-
  "https://goo.gl/bUeMhV"
```

@@

2. We concatenate the working directory file path to the desired file name for the CSV using a *function* called `paste0`. This is stored in another *object* called `student_reponses_file_name`. This creates a file name with a *file path* in your working directory and it saves the file in the folder that you are working in.

```{r, eval = FALSE}
student_responses_file_name <-
  str_c(getwd(), "./data/student-responses-data.csv")
```

@@

@@

First, let's load the data we downloaded in the last step.

```{r, message = FALSE}
student_responses <- read_csv("data/student-responses-data.csv")
```

We're also going to introduce a powerful, unusual *operator* in R, the pipe. The pipe is this symbol: `%>%`. It lets you *compose* functions. It does this by passing the output of one function to the next.

Select reduces the number of *columns* of a dataset.

```{r, eval = FALSE}
student_mot_vars <- # save object student_mot_vars by...
  student_responses %>% # using dataframe student_responses
  select(SCIEEFF, JOYSCIE, INTBRSCI, EPIST, INSTSCIE) # and selecting only these five variables

student_mot_vars
```

Also, check out the helper functions: `contains()`, `starts_with()`, and `ends_with()`, i.e.:

```{r, eval = FALSE}
student_responses %>% 
  select(contains("sci"))
```

@@

We can also rename the variables at the same time we select them. I put these on separate lines so I could add the comment, but you could do this all in the same line, too. It does not make a difference in terms of how `select()` will work. We are using `head()` to only print a few rows.

```{r}
student_responses %>% # using dataframe student_responses
  select(student_efficacy = SCIEEFF,
         student_joy = JOYSCIE, 
         student_broad_interest = INTBRSCI,
         student_epistemic_beliefs = EPIST,
         student_instrumental_motivation = INSTSCIE,
  ) %>% 
  head(3)
```

@@

# Creating a new variable

`mutate()` takes instructions to create a new variable.

What goes *before* the equals sign is the name of the new variable.

What goes after are the instructions.

```{r}
student_responses %>% 
  select(student_efficacy = SCIEEFF, 
         student_joy = JOYSCIE, 
         student_broad_interest = INTBRSCI, 
         student_epistemic_beliefs = EPIST,
         student_instrumental_motivation = INSTSCIE) %>% 
  mutate(student_joy_interest = (student_joy + student_broad_interest) / 2) %>% 
  head(3)
```

@@

# Filtering the data set

Filter takes *logical statements* (statements that can evalute to true or false) to select a number of *rows* from a dataset.

```{r}
student_responses %>% 
  select(student_efficacy = SCIEEFF, 
         student_joy = JOYSCIE, 
         student_broad_interest = INTBRSCI, 
         student_epistemic_beliefs = EPIST,
         student_instrumental_motivation = INSTSCIE) %>% 
  mutate(student_joy_interest = (student_joy + student_broad_interest) / 2) %>% 
  filter(student_joy_interest > mean(student_joy_interest, na.rm = TRUE)) %>% 
  head(3)
```

@@

# Grouping and summarizing

It is often useful to aggregate a data set.

The `group_by()` and `summarize()` functions are especially helpful for this purpose.

Let's find the mean and standard deviation (and counts) of student broad interest by teacher.

```{r}
student_responses %>% 
  select(student_broad_interest = INTBRSCI, teacher_id = SCHID) %>% 
  group_by(teacher_id) %>% 
  summarize(student_broad_interest_mean = mean(student_broad_interest, na.rm = TRUE),
            student_broad_interest_sd = sd(student_broad_interest, na.rm = TRUE),
            n = n()) %>% 
  head(3)
```

@@

# Arranging

Finally, let's arrange the table by the number of students in each class.

`arrange()` sorts in order from the smallest to largest values.

`desc()` tells `arrange()` to sort in descending order.

```{r}
student_responses %>% 
  select(student_broad_interest = INTBRSCI, teacher_id = SCHID) %>% 
  group_by(teacher_id) %>% 
  summarize(student_broad_interest_mean = mean(student_broad_interest, na.rm = TRUE),
            student_broad_interest_sd = sd(student_broad_interest, na.rm = TRUE),
            n = n()) %>% 
  arrange(desc(n)) %>% 
  head(3)
```

# Counting

Sometimes, we simply want to count the number of observations associated with each group.

```{r, eval = FALSE}
student_responses %>% 
  select(teacher_id = SCHID) %>% 
  count(teacher_id)
```

We could then use this as the basis of *other* summary statistics.

```{r}
student_responses %>% 
  select(teacher_id = SCHID) %>% 
  count(teacher_id) %>% 
  summarize(n_mean = mean(n),
            n_sd = sd(n))
```

On average, there are around 32 (*SD* = 8.61) students in each teachers' class.

class: inverse, center, middle
# Demo doc (part 1): Try it out!

# Loading data

Again, we'll load (or have loaded, already) the tidyverse library.

Just as a recap of a line that we ran earlier.

```{r, eval = TRUE, message = FALSE}
student_responses <- read_csv("data/student-responses-data.csv")
```

You can also print the name of the object.

# Loading Excel and SAV files

## Loading Excel files

## Loading SAV files

# Saving (Writing) Files

# Visualizing data

*ggplot2* is a tidyverse package for creating visualizations (or figures).

It is very flexible ... but can also be a bit challenging to use, at first.

Let's work with a smaller data set, for now.

```{r}
student_mot_vars <- # save object student_mot_vars by...
  student_responses %>% # using dataframe student_responses
  select(student_efficacy = SCIEEFF, # selecting variable SCIEEFF and renaming to student_efficiency
         student_joy = JOYSCIE, # selecting variable JOYSCIE and renaming to student_joy
         student_broad_interest = INTBRSCI, # selecting variable INTBRSCI and renaming to student_broad_interest
         student_epistemic_beliefs = EPIST, # selecting variable EPIST and renaming to student_epistemic_beliefs
         student_instrumental_motivation = INSTSCIE,  # selecting variable INSTSCIE and renaming to student_instrumental_motivation
         teacher_id = SCHID
  )
```

# Scatter plot

```{r, warning = FALSE}
ggplot(student_mot_vars,
       aes(x = student_efficacy, y = student_broad_interest)) +
  geom_point()
```

# Scatter plot with a regression line

```{r, warning = FALSE}
ggplot(student_mot_vars,
       aes(x = student_efficacy, y = student_broad_interest)) +
  geom_point() +
  geom_smooth(method = "lm")
```

# Scatter plot with a regression line and some cleaning up

```{r, warning = FALSE}
ggplot(student_mot_vars,
       aes(x = student_efficacy, y = student_broad_interest)) +
  geom_point() +
  geom_smooth(method = "lm") +
  theme_bw() +
  xlab("Student Efficacy") +
  ylab("Student Broad Interest")
```

# Scatter plot with a regression line and some cleaning up

```{r, warning = FALSE}
ggplot(student_mot_vars,
       aes(x = student_efficacy, y = student_broad_interest)) +
  geom_point() +
  geom_smooth(method = "lm") +
  theme_bw() +
  xlab("Student Efficacy") +
  ylab("Student Broad Interest")
```

class: inverse, center, middle
# Demo doc (part 2)

@@

class: inverse, center, middle
# Demo doc (part 3)

class: inverse, center, middle
# Questions, other

@@
